---
title: "Introduction to the Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The purpose of this package is to process and statistically analyze proteomics data generated from three different software platforms for raw Mass Spectrometry analysis: Proteome Discoverer, DIA-NN, and MaxQuant.  The package is designed to be user-friendly, enabling users to perform an instant, comprehensive analysis, with minimal input. The package includes four main functions that correspond to the different software specific output formats. With just one function run they complete the analysis. 

The data processing workflow includes options for:

1.	Normalization of proteomic intensity values.

2.	Filtering based on the percentage of missing values.

3.	Imputation of missing data to ensure robust downstream analysis.

Once the processing is complete, the package performs statistical analysis to identify significant differences between experimental groups. The results are automatically exported as Excel files, and a range of visualizations is generated to facilitate interpretation. These include:

•	Principal Component Analysis (PCA) plots for dimensionality reduction and group comparison.

•	Heatmap highlighting significant proteins.

•	Protein rank-abundance scatterplots.

•	Boxplots and violin plots to display data distribution and variability across groups.

The package streamlines the analysis of proteomics data, providing both detailed results and high-quality visualizations in a quick and reproducible manner.

# How to install: 

Because some package-dependencies (limma, UniProt.ws, missRanger) are inside the BioConductor repository, if they have not already been installed in your environment, use the reporting code to install the package:
```{r, eval=FALSE}
install.packages("BiocManager")
options(repos = BiocManager::repositories())
devtools::install_github("theomargel/PACKAGE")
```

Then load its library with:

```{r setup}
library(PACKAGE)
```

## Output Directory

The results from each function are saved in a folder named **MS_Analysis**, which is created inside the last directory of the provided Excel file(s).

